---
output:
  pdf_document: default
  html_document: default
---
# An Election Year  During A Politicized Global Pandemic

**Kevin Joseph, Jim Lee, Peter Kennedy, Evan O'Keefe**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warnings=FALSE)
library(tidyverse)
library(lubridate)
library(viridis)
source("ggprob.R")
```

```{r data-import, include=FALSE}
data <- read_csv("county_statistics_2.csv")
absentee_16 <- read_csv("Absentee_Ballot_Report_11_14_16.csv")
absentee_20 <- read_csv("AbsenteeCounts_County_11_6_20.csv")
```

```{r data-transforation, include=FALSE}
absentee <- absentee_20 %>%
  cbind(absentee_16) %>% 
  mutate(county = str_to_title(str_replace(Jurisdiction, " COUNTY", "")),
        returned_16 = BallotReturnedCount,
        returned_20 = BallotsReturned) %>% 
  select(county, returned_16, returned_20)

wi_data <- data %>%
  filter(state == "WI") %>% 
  left_join(absentee) %>% 
  mutate(cases_thou = cases / TotalPop * 100000,
         deaths_thou = deaths / TotalPop * 100000,
         pct_absentee_16 = returned_16 / total_votes16 * 100,
         pct_absentee_20 = returned_20 / total_votes20 * 100,
         pct_absentee_diff = (returned_20 - returned_16) / returned_16 * 100,
         third_party_16 = total_votes16 - votes16_Donald_Trump - votes16_Hillary_Clinton,
         third_party_20 = total_votes20 - votes20_Donald_Trump - votes20_Joe_Biden,
         win_margin_16 = votes16_Donald_Trump - votes16_Hillary_Clinton,
         win_margin_20 = votes20_Joe_Biden - votes20_Donald_Trump) %>% 
  drop_na()
```

```{r functions, include=FALSE}
corr_coef <- function(x, y) {
  corr <- round(cor(x, y), 2)
  corr_str <- str_c("R: ", toString(corr))
  return(corr_str)
}
```

### Introduction

2020 has been far from a normal year and two main focuses within the United States were the Coronavirus Pandemic and the Presidential Election between incumbent Republican Donald Trump and Democratic Candiate Joe Biden. The goal of our group was to analyze how these two major events impacted each other specifically in the state of Wisconsin by observing a few different factors. A key aspect that we plan to observe is the impact of increased mail-in/absentee voting. To achieve this analysis, data was collected describing Wisconsin election data (including vote totals and absentee ballot data), COVID-19 cases data (including number of cases and deaths), and general census information. This project will test the hypothesis that conservative areas of Wisconsin will have higher COVID-19 infection rates and lower absentee voting due to the general rhetoric and actions of the American Republican Party.

### Background

The datasets collected for this project contained many useful population observations; however, do to the  scope of this project only a select few were used. These variables are described below:

**County Statistics Dataset**

1. County: (str) Name of the county for which the data  was  collected.
2. State: (str) Two leter intials for each  US state. Only data from Wisconsin (WI) was used in this project.
3. percentage16_Donald_Trump: (float) Percent of the county vote for Donald Trump in the given election cycle. Given in the decimal form, maximum values of 1.
4. percentage16_Hillary_Clinton: (float) Percent of the county vote for Hillary Clinton in the given year. Given in the decimal form, maximum values of 1.
5. total_votes16: (int) Total number of county votes for the given election cycle. Includes democratic, republican, and any third party votes.
6. votes16_Donald_Trump: (int) Votes cast for Donald Trump in the given election cycle.
7. votes16_Hillary_Clinton: (int) Votes cast for Hillary Clinton in the given election cycle.
8. percentage20_Donald_Trump: (float) Percent of the county vote for Donald Trump in the given year. Given in the decimal form, maximum values of 1.
9. percentage20_Joe_Biden: (float) Percent of the county vote for Joe Biden in the given year. Given in the decimal form, maximum values of 1.
10. total_votes20: (int) Total number of county votes for the given election cycle. Includes democratic, republican, and any third party votes.
11. votes20_Donald_Trump: (int) Votes cast for Donald Trump in the given election cycle.
12. votes20_Joe_Biden: (int) Votes cast for Joe Biden in the given election cycle.
13. cases: (int) Number of COVID-19 cases in the  county as of the dataset being published.
14. deaths: (int) Number of COVID-19 deaths in the  county as of the dataset being published.
15. TotalPop: (int) Total population of the county including those not eligible to vote,
16. VotingAgeCitizen: (int) Number of citizens in the county that are eligible to vote in the election.
17. Unemployment: (float) Percent of the county that is not currently employed. Given in the percent form, maximum values of 100.

**Absentee Ballot Report 2016**

18. County: Same as previous *County*
19. BallotReturnedCount: (int) The number of absentee/mail-in ballots returned and counted during the election cycle.

**Absentee Ballot Report 2020**

20. Jurisdiction: Same as previous *County*
21. BallotsReturned: Same as previous *BallotReturnedCount*

**Calculated Values**

22. cases_thou: (int) Number of COVID-19 cases in the given county, standardized to the county population.
23. deaths_thou: (int) Number of COVID-19 deaths in the given county, standardized to the county population.
24. pct_absentee_16: (float) Percent of the county votes that come from absentee/mail-in ballots for  the  given election cycle. Given in the percent form, maximum values of 100.
25. pct_absentee_20 = Same as *pct_absentee_16*
26. pct_absentee_diff: (float) Percentage change of absentee votes from 2016 to 2020. Given in the percent form, no maximum value.
27. third_party_16: (int) Number of votes cast for the third part canidate in the given election cycle.
28. third_party_20: Same as *third_party_16*
29. win_margin_16: (int) Difference in number of votes between the winning canidate and the  losing.
30. win_margin_20: Same as *win_margin_16*

```{r repub-voter-corr, echo=FALSE}
ggplot(wi_data, aes(percentage16_Donald_Trump, percentage20_Donald_Trump)) +
  geom_point(aes(size = VotingAgeCitizen)) +
  geom_smooth(method = lm, se = FALSE) +
  labs(
    title = "Percent Republican Vote 2016 vs 2020",
    subtitle = corr_coef(wi_data$percentage16_Donald_Trump, wi_data$percentage20_Donald_Trump),
    x = "Percent Republican Votes 2016",
    y = "Percent Republican Votes 2020",
    size = "Num. Eligible Voters" 
  )
```

```{r dem-absentee-corr, echo=FALSE}
ggplot(wi_data, mapping = aes(pct_absentee_20, percentage20_Joe_Biden)) +
  geom_point(aes(size = VotingAgeCitizen)) +
  geom_smooth(method = lm, se = FALSE) +
  labs(
    title = "Percent Democratic Vote 2020 vs Absentee Vote",
    subtitle = corr_coef(wi_data$pct_absentee_20, wi_data$percentage20_Joe_Biden),
    x = "Percent Absentee Vote",
    y = "Percent Democratic Votes 2020",
    size = "Num. Eligible Voters" 
  )
```

```{r dem-cases-corr, echo=FALSE}
ggplot(wi_data, aes(percentage20_Joe_Biden, cases_thou)) +
  geom_point(aes(size = VotingAgeCitizen)) +
  geom_smooth(method = lm, se = FALSE) +
  labs(
    title = "Percent Democratic Vote 2020 vs COVID-19 Cases",
    subtitle = corr_coef(wi_data$percentage20_Joe_Biden, wi_data$cases_thou),
    x = "Percent Democratic Votes 2020",
    y = "COVID-19 Cases per 100,000",
    size = "Num. Eligible Voters" 
  )
```

```{r dem-deaths-corr, echo=FALSE}
ggplot(wi_data, aes(percentage20_Joe_Biden, deaths_thou)) +
  geom_point(aes(size = VotingAgeCitizen)) +
  geom_smooth(method = lm, se = FALSE) +
  labs(
    title = "Percent Democratic Vote 2020 vs COVID-19 Deaths",
    subtitle = corr_coef(wi_data$percentage20_Joe_Biden, wi_data$deaths_thou),
    x = "Percent Democratic Votes 2020",
    y = "COVID-19 Deaths per 100,000",
    size = "Num. Eligible Voters" 
  )
```

```{r unemploy-cases-corr, echo=FALSE}
ggplot(wi_data, aes(Unemployment, cases_thou)) +
  geom_point(aes(size = TotalPop)) +
  geom_smooth(method = lm, se = FALSE) +
  labs(
    title = "Unemployment vs COVID-19 Cases",
    subtitle = corr_coef(wi_data$Unemployment, wi_data$cases_thou),
    x = "Unemployment Percentage",
    y = "COVID-19 Cases per 100,000",
    size = "County Population" 
  )
```

```{r unemploy-deaths-corr, echo=FALSE}
ggplot(wi_data, aes(Unemployment, deaths_thou)) +
  geom_point(aes(size = TotalPop)) +
  geom_smooth(method = lm, se = FALSE) +
  labs(
    title = "Unemployment vs COVID-19 Deaths",
    subtitle = corr_coef(wi_data$Unemployment, wi_data$deaths_thou),
    x = "Unemployment Percentage",
    y = "COVID-19 Deaths per 100,000",
    size = "County Population" 
  )
```

```{r absentee-cases-corr, echo=FALSE}
no_sauk <- wi_data %>% 
  #Sauk County is excluded as it is an outlier at 1000% change
  filter(county != "Sauk") 

ggplot(no_sauk, aes(pct_absentee_diff, cases_thou)) +
  geom_point(aes(size = TotalPop)) +
  geom_smooth(method = lm, se = FALSE) +
  labs(
    title = "Absentee Vote Change vs COVID-19 Cases",
    subtitle = corr_coef(no_sauk$pct_absentee_diff, no_sauk$cases_thou),
    x = "Percentage Change in Absentee Voting",
    y = "COVID-19 Cases per 100,000",
    size = "County Population"
  )
```

```{r vote-total-calc, include=FALSE}
wi_data %>% 
  summarize(
    votes16 = sum(total_votes16),
    abs_16 = sum(returned_16),
    votes20 = sum(total_votes20),
    abs_20 = sum(returned_20)
  )

year <- c(rep("2016", 2), rep("2020", 2))
meas <- rep(c("Total Votes", "Absentee Votes"), 2)
value <- c(2893526, 820666, 3239923, 1929743)
votes <- data.frame(year, meas, value)
```

```{r vote-total-bar, echo=FALSE}
ggplot(votes, aes(year, value/100000, fill = meas)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_viridis_d(option = 'cividis') +
  labs(
    title = "Total Votes vs Number of Absentee Votes",
    x = "Election Year",
    y = "Number of Votes (Hundred Thousands)",
    fill = ""
  )
```

```{r third-party-calc, include=FALSE}
wi_data %>% 
  summarize(
    margin_16 = sum(win_margin_16),
    third_16 = sum(third_party_16),
    margin_20 = sum(win_margin_20),
    third_20 = sum(third_party_20)
  )

year <- c(rep("2016", 2), rep("2020", 2))
meas <- rep(c("Margin", "Third Party Vote"), 2)
value <- c(13604, 150284, 35712, 55573)
third_party <- data.frame(year, meas, value)
```

```{r third-party-bar, echo=FALSE}
ggplot(third_party, aes(year, value/1000, fill = meas)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_viridis_d(option = 'cividis') +
  labs(
    title = "Number of 3rd Party Voters vs Winning Margin",
    x = "Election Year",
    y = "Number of Votes (Thousands)",
    fill = ""
  )
```



